### 2014-10-30 motivation.tex
Added reference to Fig. 3 in Sec 3.

A key novelty of \name is that it allows the addition of new RDTs to the
store, greatly improving programmability.  Specifically, programmers no
longer have to coerce application logic to conform to to store-provided data
types. In addition, \name treats the convergence semantics of the data type
separately from consistency properties. This separation of concerns allows
the programmer to \emph{operationally} reason about conflict resolution,
while permitting \emph{declarative} reasoning for consistency. As mentioned
in the previous section, RDTs are described in terms of operations on a RDT
object, where each operation is described as a reduction over the state of
the object i.e., the list of effects.

Assume that the bank account object provides three operations: \cf{deposit},
\cf{withdraw} and \cf{getBalance}, with the assumption that the \cf{withdraw}
fails if the account has insufficient balance.  Every operation in \name is
of the following type, written in Haskell syntax:

@@

\noindent An operation takes a list of effects (the \emph{context} for the operation), an
argument and returns a result along with an optional effect. Read-only
operations do not return any effect. If an effect is returned, it is added to
the state of the object at the current replica, and asynchronously sent to
other replicas. The implementation of the bank account operations in \name is
given below:

@@

getBalance :: [Acc] -> () -> (Int, Maybe Acc)
getBalance ctxt _ =
  let res = sum [x | Deposit x <- ctxt]
          - sum [x | Withdraw x <- ctxt]
	in (res, Nothing)

@@

withdraw :: [Acc] -> Int -> (Bool, Maybe Acc)
withdraw ctxt v =
	if sel1 $ getBalance ctxt () >= v
  then (True, Just $ Withdraw_ v)
	else (False, Nothing)

mkOperName ''Acc
\end{codehaskell}

@@

The last line in the above code snippet uses Template Haskell~\cite{} to
generate the following datatype declaration:

\begin{codehaskell}
data OperName = Deposit | Withdraw | GetBalance
\end{codehaskell}

\noindent that declares a nullary constructor for each of the constructors in
the given type, dropping the tailing underscore from the original constructor
names. This declaration will come in handy for writing the contracts.

@@

\begin{figure}
\begin{codehaskell}
data Effect, Rel, Prop, Fol
type Contract = Effect -> Fol

Vis,So,SameObj :: Rel
(cup),(cap)      :: Rel -> Rel -> Rel
transClosure :: Rel -> Rel
((-)          :: (Effect, Effect) -> Rel -> Prop
true         :: Prop
(==),vis,so,sameObj :: Effect -> Effect -> Prop
(wedge),(vee),(=>) :: Prop -> Prop -> Prop
lift         :: Prop -> Fol
(forall) :: [OperName] -> (Effect -> Fol) -> Fol
\end{codehaskell}
\caption{Quelea contract language.}
\label{code:ctrt_lang}
\end{figure}

\name contract language is presented in Figure~\ref{code:ctrt_lang}. The
language includes primitive relations \cf{Vis},\cf{So} and \cf{SameObj},
operators to compose relations ($\cup,\cap$ and \cf{transClosure}), and check
inclusion ($\in$). The language also includes standard propositional logic
operators, and a prenex quantified $\forall$ operator. The functions \cf{vis},
\cf{so} and \cf{sameObj} are derived entities. For example, \cf{vis a b = (a,b)
$\in$ Vis}.

In our running example, it is clear that in order to preserve the integrity
constraint, the \cf{withdraw} operation must be strongly consistent. That is,
given two \cf{withdraw} operations $a$ and $b$, either $a$ is visible to $b$ or
vice versa. We express this application-level consistency as the following
contract:

\begin{codehaskell}
withdrawCtrt x = (forall) [Withdraw] $ \a -> lift $
	sameObj a x => a == e vee vis a x vee vis x a
\end{codehaskell}

\noindent Here, \cf{x} stands for the effect emitted by the current
(\cf{withdraw}) operation. The syntax \cf{\small $(\forall)
$[Withdraw]($\backslash$a $\rightarrow \ldots$)} states that $a$ is an effect
which was emitted by a \cf{withdraw} operation i.e., $\small
\oper{a}{\rcf{withdraw}}$ holds. The contract specifies that if the withdraw
operation emits an effect \cf{x}, then for any operation \cf{a} which was
emitted by a \cf{withdraw} operation, it is the case that \cf{a} and \cf{x} are
the same effect or \cf{a} is visible to \cf{x}, or vice versa. Any execution on
a bank account object that preserves the above contract for a \cf{withdraw}
operation is said to be derived from a correct implementation of \cf{withdraw}.

@@

\begin{codehaskell}
getBalanceCtrt x = (forall) [Deposit, Withdraw] $ \a ->
	(forall) [Deposit] $ \b -> (forall) [Withdraw] $ \c ->
		lift $ ((a,x) (- (So cap SameObj) => vis a x) wedge
  				 (vis b c wedge vis c x => vis b x)
\end{codehaskell}

\noindent Here, \cf{a} is an effect emitted by either a \cf{deposit} or a
\cf{withdraw} operation. The fourth line in the above snippet says that if a
\cf{deposit} or a \cf{withdraw} operation precedes a \cf{getBalance} operation
in session order, and is applied on the same object as the \cf{getBalance}
operation, then it must be the case that the \cf{getBalance} operation
witnesses the effects of that preceding operations. The last line says that if
a \cf{withdraw} $c$ is visible to \cf{getBalance}, then all \cf{deposit}
operations $b$ visible to the \cf{withdraw} should also be visible to
\cf{getBalance}. This prevents negative balance anomalies.

There are no restrictions on when and how a \cf{deposit} operation can execute.
Hence, the deposit contract imposes no consistency restrictions.

\begin{codehaskell}
depositCtrt x = lift true
\end{codehaskell}

>>>>

A key novelty	in \name is that it allows the addition of new RDTs to the store,
immensely improving the programmability of data stores. The programmer no
longer has to coerce the application logic to utilize the store provided data
types. In addition, \name treats the convergence semantics of the data type
separately from the consistency properties. The separation of concern allows
the programmer to \emph{operationally} reason about conflict resolution, while
permitting \emph{declarative} reasoning for consistency. As mentioned in the
previous section, RDTs are described in terms of operations on a RDT object,
where each operation is described as a reduction over the state of the object
i.e., the list of effects. Let us assume that the bank account object provides
three operations: \cf{deposit}, \cf{withdraw} and \cf{getBalance}, with the
assumption that the withdraw fails if the account has insufficient balance.
Every operation in \name is of the following type, written in Haskell syntax:

@@

\noindent It takes a list of effects (the \emph{context} for the operation), an
argument and returns a result along with an optional effect. Read-only
operations do not return any effect. If an effect is returned, it is added to
the state of the object at the current replica, and asynchronously sent to
other replicas. The implementation of the bank account operations in \name is
given below:

@@

getBalance :: [Acc] -> () -> (Int, Maybe Acc)
getBalance ctxt _ =
  let res = sum [x | Deposit_ x <- ctxt]
						- sum [x | Withdraw_ x <- ctxt]
	in (res, Nothing)

@@

withdraw :: [Acc] -> Int -> (Bool, Maybe Acc)
withdraw ctxt v =
	if sel1 $ getBalance ctxt () >= v
  then (True, Just $ Withdraw_ v)
	else (False, Nothing)
\end{codehaskell}

@@

@@

In our running example, it is clear that in order to preserve the integrity
constraint, the \cf{withdraw} operation must be strongly consistent.  That is,
given two \cf{withdraw} operations $a$ and $b$, either $a$ is visible to $b$ or
vice versa. Utilizing the primitive relations described in
Section~\ref{sec:sysmod}, we express this application-level consistency
requirement as a contract or specification over \cf{withdraw}:

\vspace{-1em}
\begin{smathpar}
\begin{array}{l}
\forall (a : \rcf{withdraw}).~\sameobj{a}{\cureff} \Rightarrow a = \cureff \vee \vis{a}{\cureff} \vee \vis{\cureff}{a}
\end{array}
\end{smathpar}

\noindent Here, $\cureff$ stands for the effect emitted by the \cf{withdraw} operation.
The syntax $a:\rcf{withdraw}$ states that $a$ is an effect which was emitted
by a \cf{withdraw} operation i.e., $\oper{a}{\rcf{withdraw}}$ holds.  The
contract specifies that if the current operation emits an effect $\cureff$,
then for any operation $a$ which was emitted by a \cf{withdraw} operation, it
is the case that $a = \cureff$ or $a$ is visible to $\cureff$, or vice versa.
Any execution on a bank account object that preserves the above contract for a
\cf{withdraw} operation is said to be derived from a correct implementation of
\cf{withdraw}.

@@

\vspace{-1em}
\begin{smathpar}
\begin{array}{l}
\forall (a:\rcf{deposit}), (b:\rcf{withdraw}), (c: \rcf{deposit} \vee \rcf{withdraw}). \\
\qquad \vis{a}{b} \wedge \vis{b}{\cureff} \Rightarrow \vis{a}{\cureff} \\
\qquad \wedge~ (\soZ \cap \sameobjZ) (c,\cureff) \Rightarrow \vis{c}{\cureff}
\end{array}
\end{smathpar}

\noindent The expression $c:\rcf{deposit} \vee \rcf{withdraw}$ states that $c$ is an
effect that was emitted either by a \cf{deposit} or a \cf{withdraw} operation.
If a \cf{withdraw} $b$ is visible to \cf{getBalance}, then all \cf{deposit}
operations $a$ visible to the \cf{withdraw} should also be visible to
\cf{getBalance}. This prevents negative balance anomalies.

Our specification language (Fig~\ref{fig:contract-lang}) provides
operators to compose relations. Thus, the syntax $(R_1 \cap R_2)(a,b)$
is equivalent to $R_1(a,b) \wedge R_2(a,b)$. The last line of the
above contract says that if a \cf{deposit} or a \cf{withdraw}
operation precedes a \cf{getBalance} operation in session order, and
is applied on the same object as the \cf{getBalance} operation, then
it must be the case that the \cf{getBalance} operation witnesses the
effects of the preceding operations.

Since there are no restrictions on when and how a \cf{deposit}
operation can execute, its contract is simply $\small \true$.
